---
title: "Logistic Regression analysis using Wages dataset"
output:
  word_document: default
  pdf_document: default
---

```{r, ref.label=knitr::all_labels(), echo=TRUE, eval=FALSE,include=FALSE}
#install.packages("mnormt")
#install.packages("kableExtra")
#install.packages("Matrix")
#install.packages("lme4")
#install.packages("ICC")
library(MASS)
library(gridExtra)  
library(mnormt) 
library(lme4) 
library(knitr) 
library(kableExtra)
library(tidyverse)
library(ICC)
library(ggplot2)
```


```{r, ref.label=knitr::all_labels(), echo=TRUE, eval=FALSE}
# Load the data
wagedata_raw <- read.table("C:/Users/ipseeta/Downloads/wage.data.txt", header=TRUE)

# Change variable structure
wagedata <- data.frame(as.numeric(wagedata_raw$education), as.factor(wagedata_raw$south), 
                       as.factor(wagedata_raw$sex), as.numeric(wagedata_raw$experience), 
                       as.factor(wagedata_raw$union), as.numeric(wagedata_raw$wage),
                       as.numeric(wagedata_raw$age), as.factor(wagedata_raw$race), 
                       as.factor(wagedata_raw$occupation), as.factor(wagedata_raw$sector), 
                       as.factor(wagedata_raw$marr) )

colnames(wagedata) <- c("education", "south", "sex", "experience", "union", 
                      "wage", "age", "race", "occupation", "sector", "marr")

# Check for any missing data points
any(is.na(wagedata))  # check for missing values - none

# Change wage variable to binary (<7.78=low, >7.78=high)
wagedata$wage_binary <- ifelse(wagedata$wage < 7.78, 0,1)  # change wage variable to binary high/low


head(wagedata)
str(wagedata)
```


```{r, ref.label=knitr::all_labels(), echo=TRUE, eval=FALSE}

### Proportions
par(mar = c(2, 2, 2, 2))
par(mfrow = c(4,2))

# South
south_proportion <- mean(wagedata$south == 1)
non_south_proportion <- mean(wagedata$south == 0)
barplot(c(south=south_proportion, non_south=non_south_proportion), col = c("blue", "red"), main = "Proportion in the South")
#cat("Proportion of people in south:", south_proportion, "\n")
#cat("Proportion of people not in south:", non_south_proportion, "\n")

# Sex
male_proportion <- mean(wagedata$sex == 0)
female_proportion <- mean(wagedata$sex == 1)
barplot(c(male=male_proportion, female=female_proportion), col = c("blue", "red"), main = "Proportion of Males and Females")
#cat("Proportion of males:", male_proportion, "\n")
#cat("Proportion of females:", female_proportion, "\n")

# Union
union_proportion <- mean(wagedata$union == 1)
non_union_proportion <- mean(wagedata$union == 0)
barplot(c(union=union_proportion, non_union=non_union_proportion), col = c("blue", "red"), main = "Union Membership")
# cat("Proportion of people in union:", union_proportion, "\n")
# cat("Proportion of people not in union:", non_union_proportion, "\n")



# Race
cauc_proportion <- mean(wagedata$race == 3)
hispanic_proportion <- mean(wagedata$race == 2)
other_proportion <- mean(wagedata$race == 1)
barplot(c(Caucasian=cauc_proportion, Hispanic=hispanic_proportion, Other=other_proportion), col = c("blue", "red", "green"), main = "Proportion of Race")
# cat("Proportion of caucasian:", cauc_proportion, "\n")
# cat("Proportion of hispanic:", hispanic_proportion, "\n")
# cat("Proportion of other:", other_proportion, "\n")


# Occupation
worker_proportion <- mean(wagedata$occupation == 6)
technical_proportion <- mean(wagedata$occupation == 5)
services_proportion <- mean(wagedata$occupation == 4)
office_proportion <- mean(wagedata$occupation == 3)
sales_proportion <- mean(wagedata$occupation == 2)
management_proportion <- mean(wagedata$occupation == 1)
barplot(c(Workers=worker_proportion, Technical=technical_proportion, Services=services_proportion, 
      Office=office_proportion, Sales=sales_proportion, Management=management_proportion),
    col = c("blue", "red", "green", "yellow", "orange", "purple"), main = "Proportion of each occupation")
# cat("Proportion of workers:", worker_proportion, "\n")
# cat("Proportion of technical:", technical_proportion, "\n")
# cat("Proportion of services:", services_proportion, "\n")
# cat("Proportion of office:", office_proportion, "\n")
# cat("Proportion of sales:", sales_proportion, "\n")
# cat("Proportion of management:", management_proportion, "\n")

# Sector
manufacturing_proportion <- mean(wagedata$sector == 1)
construction_proportion <- mean(wagedata$sector == 2)
other_sector_proportion <- mean(wagedata$sector == 0)
barplot(c(Manufacturing=manufacturing_proportion, Construction=construction_proportion, Other=other_sector_proportion), col = c("blue", "red", "green"), main = "Proportion of each sector")
# cat("Proportion in manufacturing sector:", manufacturing_proportion, "\n")
# cat("Proportion in construction sector:", construction_proportion, "\n")
# cat("Proportion in other sector:", other_sector_proportion, "\n")


# Married
married_proportion <- mean(wagedata$marr == 1)
not_married_proportion <- mean(wagedata$marr == 0)
barplot(c(Married=married_proportion, Not_married=not_married_proportion), col = c("blue", "red"), main = "Proportion Married")
# cat("Proportion of married people:", married_proportion, "\n")
# cat("Proportion of not married people:", not_married_proportion, "\n")

# Wage_binary
high_proportion <- mean(wagedata$wage_binary == 1)
low_proportion <- mean(wagedata$wage_binary == 0)
barplot(c(High_wage=high_proportion, Low_wage=low_proportion), col = c("blue", "red"), main = "Proportion of High vs Low wage")
# cat("Proportion of high wage:", high_proportion, "\n")
# cat("Proportion of low wage:", low_proportion, "\n")


```
```{r, ref.label=knitr::all_labels(), echo=TRUE, eval=FALSE}
#histograms for individual variables

plot1<-ggplot(wagedata, aes(education))+
  geom_histogram()

plot2<-ggplot(wagedata, aes(experience))+
  geom_histogram()

plot3<-ggplot(wagedata, aes(age))+
  geom_histogram(bins = 20)+coord_cartesian(xlim =c(10,70))

plot4<-ggplot(wagedata, aes(wage))+
  geom_histogram()
  grid.arrange(plot1,plot2,plot3,plot4)
```
```{r, ref.label=knitr::all_labels(), echo=TRUE, eval=FALSE}
created_boxplot<- boxplot(wagedata$wage)

outliers <- created_boxplot$out

wagedata<- wagedata[-which(wagedata$wage %in% outliers),]

```

```{r, ref.label=knitr::all_labels(), echo=TRUE, eval=FALSE}
# Summary of the variables


summary(wagedata)

boxplot(education~wage_binary, data=wagedata, names=c('low', 'high'), main="Education vs Wage")
boxplot(experience~wage_binary, data=wagedata, names=c('low', 'high'), main="Experience vs Wage")
boxplot(age~wage_binary, data=wagedata, names=c('low', 'high'), main="Age vs Wage")

ggplot(data=wagedata, aes(x=age, y=wage))+geom_point()

plot.11<-ggplot(data=wagedata, aes(x=education, y=wage), main="Education vs Wage (Numerical)")+geom_point()

boxplot(wage~south, data=wagedata, main="Wage vs South")
boxplot(wage~sex, data=wagedata, main="Wage vs Sex")
boxplot(wage~union, data=wagedata, main="Wage vs Union")
par(mfrow=c(1,2))
plot.12<-boxplot(wage~race, data=wagedata, main="Wage vs Race", names=c("cauc", "hispanic", "other"))
plot.13<-boxplot(wage~occupation, data=wagedata, main="Wage vs Occupation", 
        names=c("worker", "technical", "services", "office", "sales", "management") )
boxplot(wage~sector, data=wagedata, main="Wage vs Sector", names=c("manufacturing","construction","other"))
boxplot(wage~marr, data=wagedata, main="Wage vs Marr", names=c("married", "not married"))

plot.11
```
```{r, ref.label=knitr::all_labels(), echo=TRUE, eval=FALSE}
cor<-cor(wagedata[,c(1,4,6,7)])
cor
```


```{r,ref.label=knitr::all_labels(), echo=TRUE, eval=FALSE}
#removing outliers
created_boxplot<- boxplot(wagedata$education)

outliers <- created_boxplot$out

new_wagedata <- wagedata[-which(wagedata$education %in% outliers),]
dim(new_wagedata)

```

```{r, ref.label=knitr::all_labels(), echo=TRUE, eval=FALSE}
set.seed(2024)
indices<- sample(1:nrow(new_wagedata), ceiling(0.7*nrow(new_wagedata)))
train_wage<-new_wagedata[indices,]
test_wage<-new_wagedata[-indices,]
dim(train_wage)
dim(test_wage)

```

```{r, ref.label=knitr::all_labels(), echo=TRUE, eval=FALSE}
# model 1: full model
full_model <- glm(wage_binary ~ education+south+sex+experience+union+age+race+occupation+sector+marr,
                  data=train_wage, family="binomial")
summary(full_model)


# model 2: remove experience because it is highly correlated with age
model2 <- glm(wage_binary ~ education+south+sex+union+age+race+occupation+sector+marr,
                  data=train_wage, family="binomial")
summary(model2)

#drop-in-deviance for full model and model 2
anova(model2, full_model, test = "Chisq")

# model 3: quasibinomial
model3 <- glm(wage_binary ~ education+south+sex+union+age+race+occupation+sector+marr,
                  data=train_wage, family="quasibinomial")

summary(model3)




# model 4: model with all interaction terms
model4 <- glm(wage_binary ~ (education+south+sex+union+age+race+occupation+sector+marr)^2,
                  data=train_wage, family="binomial")

summary(model4)


# model 5: model with one interaction term, education*south
model5 <- glm(wage_binary ~ education+south+sex+union+age+race+occupation+sector+marr + education*south,
                  data=train_wage, family="binomial")
summary(model5)


# model 6: model with one interaction term, education*sex
model6 <- glm(wage_binary ~ education+south+sex+union+age+race+occupation+sector+marr + education*sex,
                  data=train_wage, family="binomial")
summary(model6)

#model7: interaction in south*occupation
model7<-glm(wage_binary ~ education+south+sex+union+age+race+occupation+sector+marr + south*occupation,
                  data=train_wage, family="binomial")
summary(model7)

#model8: interaction with sex and south
model8<-glm(wage_binary ~ education+south+sex+union+age+race+occupation+sector+marr + sex*south,
                  data=train_wage, family="binomial")
summary(model8)

#model9: interaction with education and south

model9<-glm(wage_binary ~ education+south+sex+union+age+race+occupation+sector+marr + education*south,
                  data=train_wage, family="binomial")
summary(model9)

#model10: interaction with south*sex(final model)
model10<-glm(wage_binary ~ education+south+sex+union+age+race+occupation+sector+marr + occupation*sex,
                  data=train_wage, family="binomial")
summary(model10)

# model11: model with two interaction term, education*south and education*sex
model11 <- glm(wage_binary ~ education+south+sex+union+age+race+occupation+sector+marr 
              + occupation*sex + education*sex, data=train_wage, family="binomial")
summary(model11)

#drop-in-deviance: model with interaction of sex and occupation and model with sex*occupation, education*sex
anova(model10, model11, test ="Chisq")

# model with numerical response variable
model12 <- glm(wage ~ education+south+sex+union+age+race+occupation+sector+marr,
                  data=train_wage)
summary(model12)


#good-ness of fit: - it indicates that the model fit the data very well.
1-pchisq(model10$deviance, model10$df.residual)
1-pchisq(model11$deviance, model11$df.residual)

```

Age and experience are highly correlated. this is because the experience is calculated using age - education

once the experience variable is removed, the age becomes a lot more statistically significant.

do not need to use an offset - we cannot use offset in logistic regression

overdispersion - when doing the quasibinomial mode, the dispersion parameter is 1.031591, which tells us that the overdispersion is very minor. thus, dont need to use quasibinomial


after testing full model including all interaction terms, none of the interactions were significant. thus, do not need to include interaction effects in the final model
the interaction effects that seemed to be significant were education&south and education&sex



the residual plots do not satisfy the assumptions because our response variable is binary. because of this, we cannot check the randomness

the plot for wage and age 


```{r,ref.label=knitr::all_labels(), echo=TRUE, eval=FALSE}
#To check the linearity of log(odds) with predictors

pred.logit<-predict(model10, type= "response")
logit<-log(pred.logit/(1-pred.logit))

#library(ggplot2)
plot1<-ggplot(train_wage, aes(train_wage$education, logit))+
  geom_point()+
  geom_smooth(method="lm", col="red")

plot2<-ggplot(train_wage, aes(train_wage$age, logit))+
  geom_point()+
  geom_smooth(method="lm", col="red")

plot3<-ggplot(train_wage, aes(train_wage$experience, logit))+
  geom_point()+
  geom_smooth(method="lm", col="red")

grid.arrange(plot1, plot2, plot3, ncol= 3)
```


```{r, ref.label=knitr::all_labels(), echo=TRUE, eval=FALSE}

pred.test<-predict(model10, newdata= test_wage, type= "response")
label_test<-ifelse(pred.test>=0.5, "1", "0")
mean(label_test!= test_wage$wage_binary)
mean(label_test==test_wage$wage_binary)
```

#LASSO model

```{r, ref.label=knitr::all_labels(), echo=TRUE, eval=FALSE}
set.seed(1)

library(glmnet)

x <- model.matrix(wage_binary~., train_wage[,-c(4,6)])
y <- train_wage[,12]

#fitting model using different lambda values
grid <- 10^seq(10, -2, length=100)
lasso.mod.wage <- glmnet(x, y, alpha=1, lambda=grid, family="binomial")

#using 10-fold cv to select optimal lambda
cv.fit.lasso.wage <- cv.glmnet(as.matrix(x), as.matrix(y), alpha=1, family="binomial")

cv.fit.lasso.wage$lambda.min
#using optimal lambda to fit the LASSO regression
lasso.mod_best.wage <- glmnet(x, y, alpha=1, lambda=0.007389478, family="binomial")
summary(lasso.mod_best.wage)
lasso.mod_best.wage
coef(lasso.mod_best.wage)

#plots for lambda and norm
library(plotmo)
plot(cv.fit.lasso.wage)
par(mfrow= c(1,2))
plot_glmnet(cv.fit.lasso.wage$glmnet.fit, s=0.007389478, label =12, xvar = "norm", xlab ="L1-norm")
plot_glmnet(lasso.mod.wage, xvar= "rlambda",s=0.007389478, xlim=c(5,-5))

test_x <- model.matrix(wage_binary~., test_wage[,-c(4,6)])
test_y <- test_wage[,12]

#predicting values
preds_lasso.wage<-predict(lasso.mod_best.wage, test_x, type = "response")
preds_label_lasso.wage<-ifelse(preds_lasso.wage>0.5, "1","0")
mean(preds_label_lasso.wage!=test_y)
mean(preds_label_lasso.wage==test_y)

```




